---
title: "Depression Trend"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(plotly)
library(mgcv)
library(modelr)
library(ggmosaic)

library(lmtest)
library(performance)
library(knitr)
library(kableExtra)
library(patchwork)

library(usmap)
library(gridExtra)
library(dplyr)
library(readxl)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 9, 
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(
  theme_minimal() + 
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5)
    )  
)
```




```{r, message=FALSE}
anx_dep = 
  read_csv("data/nhis_data01.csv") %>% 
  janitor::clean_names() %>% 
  filter(year>=2015) %>% 
  select(year, worrx, worfreq, worfeelevl, deprx, depfreq, depfeelevl, age, sex, marst, poverty) %>% 
  mutate(
    sex = recode_factor(sex, 
                        "1" = "Male", 
                        "2" = "Female"),
    marst = recode_factor(marst, 
                        "10" = "Married", "11" = "Married", "12" = "Married", "13" = "Married",
                        "20" = "Widowed",
                        "30" = "Divorced",
                        "40" = "Separated",
                        "50" = "Never married"),
    poverty = recode_factor(poverty, 
                        "11" = "Less than 1.0", "12" = "Less than 1.0", 
                        "13" = "Less than 1.0", "14" = "Less than 1.0",
                        "21" = "1.0-2.0", "22" = "1.0-2.0", 
                        "23" = "1.0-2.0", "24" = "1.0-2.0", 
                        "25" = "1.0-2.0",
                        "31" = "2.0 and above","32" = "2.0 and above",
                        "33" = "2.0 and above","34" = "2.0 and above",
                        "35" = "2.0 and above","36" = "2.0 and above",
                        "37" = "2.0 and above","38" = "2.0 and above"),
    worrx = recode_factor(worrx,
                          '1' = "no", 
                          '2' = "yes"),
    worfreq = recode_factor(worfreq, 
                            '1' = "Daily", 
                            '2' = "Weekly", 
                            '3' = "Monthly", 
                            '4' = "A few times a year", 
                            '5' = "Never"),
    worfeelevl = recode_factor(worfeelevl, 
                               '1' = "A lot", 
                               '3' = "Somewhere between a little and a lot", 
                               '2' = "A little"),
    deprx = recode_factor(deprx, '1' = "no", '2' = "yes"),
    depfreq = recode_factor(depfreq, '1' = "Daily", '2' = "Weekly", 
                            '3' = "Monthly", '4' = "A few times a year", 
                            '5' = "Never"),
    depfeelevl = recode_factor(depfeelevl, '1' = "A lot", 
                               '3' = "Somewhere between a little and a lot", 
                               '2' = "A little"),
    age = ifelse(age>=85, NA, age)
    )
```




## Depression Trend

### Percentage of people reported taken medication for depression
According to the plot, the proportion of people reported taken medication for depression increased from 8.75% in 2015 to 11.42% in 2020, followed by a slight decrease from 2020 to 2021. COVID-19 appears to have a limited impact on depression.

```{r, message=FALSE}
anx_dep %>%
  drop_na(deprx) %>% 
  group_by(year, deprx) %>% 
  summarize(dep_num = n()) %>% 
  pivot_wider(
    names_from = deprx,
    values_from = dep_num
  ) %>% 
  mutate(
    dep_percentage = yes/(no + yes)*100,
    text_label = str_c(yes, " out of ", no + yes)
  ) %>% 
  ungroup() %>% 
  plot_ly(
    y = ~dep_percentage,
    x = ~year,
    color = ~year,
    type = "bar", 
    colors = "viridis",
    text = ~text_label
  ) %>% 
  layout(
    xaxis = list (title = ""),
    yaxis = list (title = "Percentage"),
    showlegend = FALSE
  ) %>% 
  hide_colorbar()
```

#### Stratify by Biological sex

Stratify the reported percentage of people taking medication for depression by biological sex, we can observe a much higher percentage among females than males. There are also a faster increase in the percentage among females from 12.68% in 2017 to 15.14% in 2020 and a decrease from 15.14% in 2020 to 14.52% in 2021. The decrease may indicate that COVID-19 can be a protective factor against depression among females. Contrary to females, the percentage slightly decreased from 2018 to 2019 and then increased from 2020 to 2021 among males.

```{r, message=FALSE}
anx_dep %>%
  drop_na(sex, deprx) %>% 
  group_by(sex, year, deprx) %>% 
  summarize(dep_num = n()) %>% 
  pivot_wider(
    names_from = deprx,
    values_from = dep_num
  ) %>% 
  mutate(
    dep_percentage = yes/(no + yes)*100,
    text_label = str_c(yes, " out of ", no + yes)
  ) %>% 
  ungroup() %>% 
  plot_ly(
    y = ~dep_percentage,
    x = ~year,
    color = ~sex,
    type = "bar", 
    colors = "viridis",
    text = ~text_label
  ) %>% 
  add_trace(
    x = ~year,
    y = ~dep_percentage,
    color = ~sex,
    type='scatter',
    mode='lines+markers'
  ) %>% 
  layout(
    xaxis = list (title = ""),
    yaxis = list (title = "Percentage"),
    legend = list(orientation = 'h')
  )

```

#### Stratify by ratio of family income to poverty threshold

Stratify the percentage of people reported taken medication for depression by the ratio of household income to the poverty line, we can clearly see that the poorer the household, the higher their percentage. The percentage among the poorest stratum decreased from 17.41% in 2017 to 16.53% in 2018, which is the opposite of what happened in the other two strata. The change in the percentage is quite stable from 2018 to 2019 among all three strata. There is a rapid increase from 17.02% in 2019 to 18.66% in 2020, which may indicate that people belonging to the poorest stratum are affected by COVID-19 related depression. From 2020 to 2021, the percentage decreases for the other two strata, while for the richest strata, the percentage increases.


```{r, message=FALSE}
anx_dep %>%
  drop_na(poverty, deprx) %>% 
  group_by(poverty, year, deprx) %>% 
  summarize(dep_num = n()) %>% 
  pivot_wider(
    names_from = deprx,
    values_from = dep_num
  ) %>% 
  mutate(
    dep_percentage = yes/(no + yes)*100,
    text_label = str_c(yes, " out of ", no + yes)
  ) %>% 
  ungroup() %>% 
  plot_ly(
    y = ~dep_percentage,
    x = ~year,
    color = ~poverty,
    type = "scatter", 
    mode = "lines+markers",
    colors = "viridis",
    text = ~text_label
  ) %>% 
  layout(
    xaxis = list (title = ""),
    yaxis = list (title = "Percentage"),
    legend = list(orientation = 'h')
  )
  
```

#### Stratify by current martial status
Stratify the percentage of people reported taken medication for depression by current martial status, we can observe a rapid decrease from 17.26% in 2016 to 13.12% in 2019 among separated, while this downward trend slows from 2018 to 2019 and reverses from 2019 to 2020. The trends are similar for married and never married, divorced and widowed. the effect of COVID-19 is not evident in this plot.

```{r, message=FALSE}
anx_dep %>%
  drop_na(marst, deprx) %>% 
  group_by(marst, year, deprx) %>% 
  summarize(dep_num = n()) %>% 
  pivot_wider(
    names_from = deprx,
    values_from = dep_num
  ) %>% 
  mutate(
    dep_percentage = yes/(no + yes)*100,
    text_label = str_c(yes, " out of ", no + yes)
  ) %>% 
  ungroup() %>% 
  plot_ly(
    y = ~dep_percentage,
    x = ~year,
    color = ~marst,
    type = "scatter",
    mode='lines+markers',
    colors = "viridis",
    text = ~text_label
  ) %>% 
  layout(
    xaxis = list (title = ""),
    yaxis = list (title = "Percentage"),
    legend = list(orientation = 'h')
  )
```


### Age distribution

As we can see from the plot, the age distribution of people taking medication for depression did not change much from 2015 to 2021.

```{r, message=FALSE}
anx_dep %>%
  drop_na(age, deprx) %>% 
  ggplot(
    aes(x=age, group=deprx, fill=deprx)
  ) +
  geom_density(alpha=0.4) +
  facet_wrap(~year) +
  labs(
    fill = "Whether taken medicine for depression"
  )
```

### Frequency of depression

From this bar plot about how often people feel depressed, we can observe that the frequency is quite stable with a decrease from 2018 to 2019 and am increase from 2019 to 2021. There is no clear evidence of an effect of COVID-19 on the frequency of depression.

```{r, message=FALSE}
anx_dep %>% 
  drop_na(depfreq) %>% 
  group_by(year, depfreq) %>% 
  summarize(count = n()) %>% 
  group_by(year) %>% 
  summarize(
     percentage=100 * count/sum(count),
     sum_count = sum(count),
     depfreq = depfreq,
     count=count
  ) %>% 
  mutate(
    text_label = str_c(count, " out of ", sum_count)
  ) %>% 
  plot_ly(
    y = ~percentage,
    x = ~year,
    color = ~depfreq,
    type = "bar", 
    colors = "viridis",
    text = ~text_label
  ) %>% 
  layout(
    xaxis = list (title = ""),
    yaxis = list (title = "Percentage"), 
    barmode = 'stack',
    legend = list(orientation = 'h')
  )
```

### Level of depression
From this bar plot about the level of depression last time, we can see that the percentage of people who felt a lot or between a little and a lot depression is stable over the time period and a decrease of percentage of people feel a lot depression from 2018 to 2019. There is also no clear evidence of an effect of COVID-19 on the level of depression.

```{r, message=FALSE}
anx_dep %>%
  drop_na(depfeelevl) %>% 
  group_by(year, depfeelevl) %>% 
  summarize(count = n()) %>% 
  group_by(year) %>% 
  summarize(
     percentage=100 * count/sum(count),
     sum_count = sum(count),
     depfeelevl = depfeelevl,
     count=count
  ) %>% 
  mutate(
    text_label = str_c(count, " out of ", sum_count)
  ) %>% 
  plot_ly(
    y = ~percentage,
    x = ~year,
    color = ~depfeelevl,
    type = "bar", 
    colors = "viridis",
    text = ~text_label
  ) %>% 
  layout(
    xaxis = list (title = ""),
    yaxis = list (title = "Percentage"), 
    barmode = 'stack',
    legend = list(orientation = 'h')
  )
```



